Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-06-27 trash/paper2.tex
Edits to Introduction

\title{Exploiting Intrinsic Noise to Control Cell Differentiation Dynamics in the Presence of Time Delays, Inexact Models, and Extrinsic Variability}
\maketitle

@@

Previous research has focused on robust control performance in the presence of noise, but understanding controllers that exploit noise remains incomplete. 
Motivated by Maxwells-Demon, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals. We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
 We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.  Together, these findings suggest that noise-exploiting control should be possible where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.

\end{abstract}
Keywords: Stochastic Control, Gene Regulation, Optogenetics

@@

\section{Introduction}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Cartoons.pdf}
\caption{{\bf Optogenetic control of multiple cells using a single light input.}
({\bf A}) Schematic of the light activated genetic system with autoregulation.
({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
({\bf C}) Adapted from (PAPER). Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III). Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}
% XXX - Please modify figure as follows: (1) switch panels A and B to match caption. (2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".  (3) Change title of panel A (previously panel B) to "Internal gene regulation pathway". (4) Change title of panel C to "Feedback control law and resulting performance"
% MMM - Made the edit

Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles\cite{XXX}, production machinery\cite{XXX}, or chemical processes\cite{XXX}.
Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \cite{XXX}.
For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem)\ref{XXX}, and most current approaches seek to control the system to minimize variations that result from the noisy inputs.

Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation\cite{XXX} or mRNA translation regulation \cite{XXX} or to modify these processes for practical use\cite{XXX}.
However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, additive noise model is far less realistic.
In this case, Brownian motion, discrete stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage. 
At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation). 
The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it, and despite these challenges, cells generally demonstrate strong capability to survive these noisy processes. 
When seeking to understand how such systems evolve or react, the central limit theorem and the Gaussian noise may not apply, and a more detailed statistical analysis of probabilistic behavior is needed uncover hidden properties of cellular control mechanisms.

One emerging field that is particularly dependent on the integration of control and noise is synthetic biology, which aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}. 
Advances in optogenetics have enhanced the ability to reliably actuate embedded systems within cells, offering the potential to exert precise temporal and spatial control on cellular components\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. 
These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} . 
Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} to control synthetic systems to different stable points. 
% XXX Missing Citation above.

While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. 
It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure\ \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

This paper extends the analysis of the SIMO multicellular control problem by examining the impact of model uncertainties and fluctuating control objectives on the control performance. 
These uncertainties include coarse-grained approximations of the system dynamics, errors or extrinsic variations in the system parameter, and time delays between the observation of the cellular dynamics and actuation of the control process. 
In the following `Methods' we introduce our formulation of the chemical master equation to analyze the discrete stochastic distribution of cellular responses; we define multiple controllers and demonstrate the computation of their effects on cell dynamics; and we show how the control law can be optimized to improve performance.
In the `Results' section, we explore the how model approximations, parameter inaccuracies, and time delays affect control performance, and we demonstrate a simple scheme for controlling cells to track a dynamically changing reference signal. 
Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to follow different desired trajectories at different frequencies and phases.

%Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.  Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions. Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.

\section{Methods}
In May et al. \cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.  These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation. These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6). The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. Both models were independently parameterized using the same experimental data from Baumschlager et al. \cite{XXX}. Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}. Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal, and we also showed that although this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model. In the current work, we adopt the simpler of the two models and extend our analyses  to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals. Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and finally Sections \ref{sec:Scaling} and \ref{sec:Delay} introduces uncertainties into the model related t the granularity of the model approximation or the introduction of time delays, respectively.

\subsection{Model}\label{sec:Model}
To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al. \cite{May2021} (refer to Fig. \ref{cartoons}A). This model comprises two reactions for production and degradation of the key protein. The nonlinear, UV-dependent production is defined by the following equation:
\begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter. 
Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
 Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
the baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021}.

\begin{table}[]
\caption{Baseline Model Parameters}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value   & units                     \\ \hline \hline
$\kappa$         & 0.406   & Molecules Per Minute \\ \hline
$\beta$          & 20.0    & Molecules            \\  \hline
$\eta $           & 8.00    & unit less          \\  \hline
$k_0$           & 0.0001  & Molecules per Minute \\  \hline
$\gamma$         & 0.0203  & per Minute         \\  \hline
$u$              & dynamic & Molecules per Minute \\ \hline
\end{tabular}
\label{table}
\end{center}
\vspace{-0.2in}
\end{table}

\subsection{Stochastic analyses of the model}\label{sec:Stochastic}
%Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
%To recast the above ODE formulation into the discrete stochastic representation, 
To describe the discrete behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$), where the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
\begin{align}
\mathbf{s}_1 &= \mathbf{e}_1,\ \mathbf{s}_2 = -\mathbf{e}_1,\nonumber\\ 
\mathbf{s}_3 &= \mathbf{e}_2,\ \mathbf{s}_4 = -\mathbf{e}_2,\nonumber\\
&\vdots \nonumber\\ 
\mathbf{s}_{\rm 2N_c-1} &= \mathbf{e}_{\rm N_c},\ \mathbf{s}_{\rm 2N_c} = -\mathbf{e}_{\rm N_c}, \label{Stochs}
 \end{align}
where each $\mathbf{e}_{i}\in\mathbb{Z}^{{\rm N_ c}}$ is is a Euclidean vector (i.e., unity for the $i^{\rm th}$ entry and otherwise zero). The corresponding propensity functions are:
\begin{align}
w_1(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2(\mathbf{X}) = \gamma x_1,\nonumber\\
w_3(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4(\mathbf{X}) = \gamma x_2,\nonumber\\
&\vdots \nonumber\\
w_{\rm 2N_c-1}(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c}(\mathbf{X}) = \gamma x_N.\label{Props}
 \end{align}

The stoichiometry and propensities allow us to use the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process. However, a direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design. The CME, based on the same stoichiometry and propensity functions, is a linear ODE that describes the time-dependent changes in probability mass due to the specified reaction propensities and stoichiometries. It can be expressed as:
\begin{equation}
\frac{d}{dt}P(\mathbf{X}_i)=\sum_{\mu=1}^{2{\rm N_c}}
\left(
-w_{\mu}(\mathbf{X}_i)P(\mathbf{X}_i)
+w_{\mu}(\mathbf{X}_i-\mathbf{s}_{\mu})P(\mathbf{X}_i-\mathbf{s}_{\mu})
\right).\label{CMEindex}
\end{equation}
For convenience, the CME can also be formulated more compactly in matrix/vector form as:
\begin{equation}
\frac{d}{dt}\mathbf{P}=(\mathbf{A}_0+\textbf{Bu}^{\mathcal{C}})\mathbf{P},\label{CME}
\end{equation}
where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the contribution that these control inputs make to the infinitesimal generator when included into the feedback process.

@@

\subsection{Quantification and optimization of control performance}\label{sec:Quantification}
%To quantify the performance of the controlled system, Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.

In the SIMO control of the multiple cell process, any input signal applied to one cell affects all cells in the population. Thus, it is not possible to independently control individual cells without impacting others. An effective controller must strike a balance among the desired behaviors of all cells in the system.

To quantify performance success, we define the {\em steady state performance error}, $J$, as the expected steady state Euclidean distance of the process from the specified target state, $\mathbf{T}$:
 \begin{equation}
 J = \mathbb{E}\{|\mathbf{X}-\mathbf{T}|_2\}.
 \end{equation}
This score is easily calculated by applying a linear operation on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
\begin{align}
J^2&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i- T_1)^2 + (j- T_2)^2 +\ldots\right],\nonumber  \\
&=\mathbf{C}\mathbf{P}^*,
\label{Euclid}
\end{align}
where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. As a result of this calculation, $J$ is a non-negative scalar that is zero if any only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.

We consider two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that bases its control signal  on simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies only on observations from a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2)
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_2)
\end{equation}
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. In \cite{May2021}, we optimized each of these controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ for a single target state $\mathbf{T} = [XXX,XXX]$, and these controllers are assumed to be fixed for the current study unless noted otherwise.

\subsection{Scaling for system granularity}\label{sec:Scaling}
%Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}. 
%
%However, using CME analyses is often hindered by an experimental limitations.  Gene expression quantification based on fluorescence intensities, such as using fluorescent protein reporters, typically captures relative changes in expression over time rather than providing direct molecular counts.
%% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
%
%It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system. This would enable the analysis of models with more or less noise without the need of new unique models each time.  A rescaling of the system could decrease the intrinsic noise while maintaining the same general behavior. 
%To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.

>>>>

\title{Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability}
\maketitle

@@

Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
Motivated by optogentics, we proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
In previous work, we found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
\end{abstract}
Keywords: Optogenetic control, Stochastic gene regulation, Synthetic biology

@@

\section{Introduction}

Excellent work has been done to make control problems robust to noise, but these control problems control their systems despite the noise rather than working with the noise.
Many control problems set up in this way have less control performance than a noise free one, but It may be possible to use nonlinearity, and feedback control to improve control.

robust control

noise exploiting controlers were developed in may.2022 which use noise to thier advantage

Maxwells demon

optogenetics and noise in optogenetics
Many biological systems are stochastic and require

master equation

SIMO control of optogentically motivated systems

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{CartoonAndController.pdf}
\caption{{\bf Optogenetic control of multiple cells using a single light input.}
({\bf A}) Schematic of the light activated genetic system with autoregulation.
({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
({\bf C}) Adapted from (PAPER). Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III). Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}

\section{Methods}
In previous work~\cite{May2021}, a six-species gene regulation model was developed to describe light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6). A second, far simpler single-species model was also developed by assuming quasi-steady equilibrium for the first five species. Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below). In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance would change upon additional simplifying assumptions or errors in model specification.  In the following subsections, we discuss our methods to formulate this model, show how it is analyzed numerically, and discuss the specification of the control objective and the determination of an optimal controller.

\subsection{Model}
To examine how model approximations would affect the application of noise-enhanced control strategies, we start with the one species model from \cite{May2021} (see Fig.\ \ref{cartoons}A), which consists of two reactions: protein production and degradation.  The nonlinear light-activated protein production rate is given by
 \begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength. Through external feedback, the level of light excitation, and therefore the T7 promoter strength, $u(UV(t))$, can be modulated in response to the state of the system, thus removing explicit dependence on time. For example, in a system with the observation of two cells, one could define $u(UV(t)) = u^{\rm (FAC)}(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the fully aware control law, and for a system with the observation of only one cell, one could define $u^{\rm (PAC)}(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the partially aware control law. Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
Parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$
%were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and 
% XXX - I don't think that this is right.  Only the parameters u(UV) and gamma were fit to data.
% MMM - yes that is right
are presented in Table 1~\cite{May2021}.

\begin{table}[]
\caption{Model Parameters}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value   & units                     \\ \hline
$\kappa$         & 0.406   & Molecules Per Minute \\
$\beta$          & 20.0    & Molecules            \\
$\eta $           & 8.00    & unit less          \\
$k_0$           & 0.0001  & Molecules per Minute \\
$\gamma$         & 0.0203  & per Minute         \\
$u$              & dynamic & Molecules per Minute \\ \hline
\end{tabular}
\label{table}
\end{center}
\vspace{-0.2in}
\end{table}

\subsection{Stochastic analyses of the model}
%Solving discrete stochastic representations of chemical dynamics is critical to the development of noise-exploiting controllers, and can be analyzed using the Stochastic Simulation Algorithm (SSA) , or the Finite State Projection (FSP).
%To recast the above ODE formulation into the discrete stochastic representation, 
To describe the discrete behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
The stoichiometry vector for a reaction in any given cell is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$), where the $2N_c$ possible reactions are defined in pairs corresponding to production (odd indices) and degradation (even indices) as:
\begin{align}
\mathbf{s}_1 &= \mathbf{e}_1,\ \mathbf{s}_2 = -\mathbf{e}_1,\nonumber\\ 
\mathbf{s}_3 &= \mathbf{e}_2,\ \mathbf{s}_4 = -\mathbf{e}_2,\nonumber\\
&\vdots \nonumber\\ 
\mathbf{s}_{\rm 2N_c-1} &= \mathbf{e}_{\rm N_c},\ \mathbf{s}_{\rm 2N_c} = -\mathbf{e}_{\rm N_c}, \label{Stochs}
 \end{align}
where each $\mathbf{e}_{i}\in\mathbb{Z}^{{\rm N_ c}}$ is is a Euclidean vector (i.e., unity for the $i^{\rm th}$ entry and otherwise zero). The corresponding propensity functions are:
\begin{align}
w_1(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2(\mathbf{X}) = \gamma x_1,\nonumber\\
w_3(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4(\mathbf{X}) = \gamma x_2,\nonumber\\
&\vdots \nonumber\\
w_{\rm 2N_c-1}(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c}(\mathbf{X}) = \gamma x_N.\label{Props}
 \end{align}

With these stoichiometry and propensities, we can formulate and run the Gillespie stochastic simulation algorithm\cite{Gillespie1992,Gillespie1977} to generate representative trajectories of the stochastic process. However, quantifying performance and designing an optimal controller requires a more direct analysis of the chemical master equation (CME).  For the same specifications of the stoichiometry and propensity functions, the CME is a linear ODE that describes how probability mass changes in time due to the specified reaction propensities and stoichiometries, which can be written:
\begin{equation}
\frac{d}{dt}P(\mathbf{X}_i)=\sum_{\mu=1}^{2{\rm N_c}}
\left(
-w_{\mu}(\mathbf{X}_i)P(\mathbf{X}_i)
+w_{\mu}(\mathbf{X}_i-\mathbf{s}_{\mu})P(\mathbf{X}_i-\mathbf{s}_{\mu})
\right).\label{CMEindex}
\end{equation}
For convenience, the CME can also be formulated more compactly in matrix/vector form as:
\begin{equation}
\frac{d}{dt}\mathbf{P}=(\mathbf{A}_0+\textbf{Bu}^{\mathcal{C}})\mathbf{P},\label{CME}
\end{equation}
where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the contribution that these control inputs make to the infinitesimal generator when included into the feedback process.

@@

\subsection{Quantification and optimization of control performance}
%To quantify the performance of the controlled system, Score metrics of steady state distributions were made in order to enable optimization of the system by adjusting controllers.

For the single-input-multiple-output control of the multiple cell system, any input signal that is applied to one cell in the population will be felt by all cells in the population. As a result, it is not possible to independently control any individual cell without introducing changes to the behavior of the others. Therefore, an effective controller must be able to strike a balance among the desired behaviors of all cells in the system.

To quantify performance success, we define the performance error score, $J$, as the expected steady state squared Euclidean distance of the process from the specified target state, $\mathbf{T}$:
 \begin{equation}
 J = E\{(\mathbf{X}-\mathbf{T})^2\}.
 \end{equation}
This score is easily calculated by applying a linear operation on steady state probability distribution $\mathbf{P}^*$ from Eq.\ \ref{SSDist}) as follows:
{ \begin{align}
J&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i-\mathbf T_1)^2 + (j-\mathbf T_2)^2 +\ldots\right],\nonumber  \\
&=\sum_{i,j,\ldots} P^*_{ij\ldots}(t)C_{ij\ldots} =\mathbf{C}\mathbf{P}^*,
\label{Euclid}
\end{align}} where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. As a result of this calculation, $J$ is a non-negative scalar that is zero if any only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.

In this work, we explore two different controller designs that allow for the controller to use two different types of information. A fully aware controller ($\mathbf{u}^{FAC}$) uses direct observation of the protein count of two cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) uses only the direct observation of a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2)
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_1)
\end{equation}
% XXX - Do we need to use 0 and 1.  Why not use 1 and 2 so that this can match better to the formulaton above?  I changed them to x1 and x2 here, but please chekc to make sure that this is sonsistent throughout text and figures.
% MMM - Made the change to x1 x2
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. In \cite{May2021}, the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized. In what follows, we will use these controllers exactly and without modification to explore how performance changes when the controllers are applied to systems that are approximations of the model under they were originally designed.

\subsection{Scaling for system granularity}
One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species. To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.
